WIP: Transparent Standby for Low-Power, Resource-Constrained Embedded Systems

### 2018-05-01 LCTES/body.tex
(>) LCTES: versão final

According to the International Energy Agency (IEA), the number of
network-connected devices is expected to reach 50 billion by 2020 with the
expansion of the Internet of Things (IoT)~\cite{iea.data}.
%
However, most of the energy to power these devices will be consumed in
\emph{standby mode}, i.e., when they are neither transmitting or processing
data.
%
For instance, standby power currently accounts for approximately 10--15\% of
residential electricity consumption, and $CO_2$ emissions related to standby
are equivalent to those of 1 million cars~\cite{iea.data,standby.australia}.
%
The projected growth of IoT devices, together with the surprising effects of
standby consumption, made network standby efficiency one of the six
pillars of the \emph{G20's Energy Efficiency Action Plan}%
\footnote{G20's Energy Efficiency Action Plan: \url{https://www.iea-4e.org/projects/g20}}.
%
However, making effective use of standby requires software-related efforts in
order to detect idle periods of activity in a device, identify peripherals
that must remain functional, and apply appropriate sleep mode levels in its
microcontroller.

Given the projected scale of the IoT, the role of low-power standby towards
energy efficiency, and the posed software-related challenges, our research has
the following goals:
%
%\begin{enumerate}
(i)   address energy efficiency through rigorous use of standby;
(ii)  target low-power, resource-constrained embedded architectures that form
      the IoT;
(iii) provide standby mechanisms at the programming language level that scale
      to all applications; and
(iv)  support transparent/non-intrusive standby mechanisms that reduce barriers
      of adoption.
%\end{enumerate}

Our proposal lies at the bottom of the software development
layers---programming language mechanisms---meaning that \emph{all} applications
should take advantage of low-power standby modes automatically, without extra
programming efforts.
%
We extend the programming language \CEU~\cite{ceu.sensys13,ceu.tecs17} with
support for interrupt service routines (ISRs) and with a simple power
management runtime (PMR).
%
In contrast with other concurrency models (e.g., thread and actor based), the
synchronous semantics of \CEU guarantees that reactions to the environment
always reach an idle state amenable to standby.
%
Previous work~\cite{ceu.sensys13} evaluates the expressiveness of \CEU in the
context of Wireless Sensor Networks and discusses the development of drivers,
network protocols, and full applications in the language.
In addition, \CEU incurs a small overhead of memory in comparison to C (around
5-10\%), thus being a suitable alternative for constrained devices.
The language runtime is in the order of a few kilobytes only (less than 5Kb).

In our approach, each supported microcontroller requires hooks in C for the
ISRs and PMR, and each peripheral requires a driver in \CEU.
These are write-once code that is typically packaged and distributed in a
software development kit (SDK).
%
Then, all new applications built on top of these drivers take advantage of
standby automatically.
%
As a proof of concept, we provide an open source SDK %
%\footnote{https://github.com/fsantanna/ceu-arduino/}
with support for 8-bit
\emph{AVR/ATmega} and 32-bit \emph{ARM/Cortex-M0} microcontrollers, and a
variety of peripherals, such as for GPIO, A/D converter, USART, SPI, and the
nRF24L01 transceiver.
%
We developed a number of simple applications using these peripherals
concurrently and could verify that the applications remain in the deepest
standby modes for the longest periods of time.

@@

In Section~\ref{sec.ceu}, we compare the structure of programs in \CEU and
Arduino~\cite{arduino.book}, whose primary goal is to reduce the programming
barrier of adoption for a non-technical audience (e.g., designers and artists).
We show that we can keep the intended sequential reasoning of Arduino even when
applications require non-trivial concurrent behavior.
%
In Section~\ref{sec.standby}, we discuss the software infrastructure that
allows for unmodified programs in \CEU to take advantage of standby
automatically.
%
In Section~\ref{sec.conclusion}, we discuss future work and conclude the paper.

@@

With all the standby infrastructure set, the unmodified program of
Figure~\ref{lst.inversion}.b will automatically take advantage of the deepest
sleep modes for the longest periods of time possible.

@@

In future work, in order to evaluate the gains in energy efficiency with the
proposed infrastructure, we will evaluate the consumption of realistic
applications.
%
The Arduino community has an abundance of open-source projects which can be
rewritten in \CEU to take advantage of transparent standby.
%
In this scenario, we can evaluate the time to rewrite, the resulting program
structure, and the actual energy efficiency.

\begin{acks}
The authors would like to thank Guilherme Simas for the initial explorations
with the language extensions and development of some of the device drivers.
\end{acks}

>>>>

% TODO-FULL: voltar c/ artigo completo
\begin{comment}
According to the International Energy Agency (IEA), the number of
network-connected devices is expected to reach 50 billion by 2020 with the
expansion of the Internet of Things (IoT)~\cite{iea.data}.
%
However, most of the energy to power these devices will be consumed in
\emph{standby mode}, i.e., when they are neither transmitting or processing
data.
%
For instance, standby power currently accounts for approximately 10--15\% of
residential electricity consumption, and $CO_2$ emissions related to standby
are equivalent to those of 1 million cars~\cite{iea.data,standby.australia}.
%
The projected growth of IoT devices, together with the surprising effects of
standby consumption, made network standby efficiency one of the six
pillars of the \emph{G20's Energy Efficiency Action Plan}%
\footnote{G20's Energy Efficiency Action Plan: \url{https://www.iea-4e.org/projects/g20}}.
%
However, making effective use of standby requires software-related efforts in
order to detect idle periods of activity in a device, identify peripherals
that must remain functional, and apply appropriate sleep mode levels in its
microcontroller.

Given the projected scale of the IoT, the role of low-power standby towards
energy efficiency, and the posed software-related challenges, our research has
the following goals:
\end{comment}
%%%%
According to the International Energy Agency (IEA), the number of
network-connected devices is expected to reach 50 billion by 2020 with the
expansion of the Internet of Things (IoT)~\cite{iea.data}.
%
Most of the energy to power these devices will be consumed in
\emph{standby mode}, i.e., when they are neither transmitting or processing
data.
%
However, making effective use of standby requires software-related efforts in
order to detect idle periods of activity in a device, identify peripherals
that must remain functional, and apply appropriate sleep mode levels in its
microcontroller.
Therefore, our research has the following goals:
%%%%
%
%\begin{enumerate}
(i)   address energy efficiency through rigorous use of standby;
(ii)  target low-power, resource-constrained embedded architectures that form
      the IoT;
(iii) provide standby mechanisms at the programming language level that scale
      to all applications; and
(iv)  support transparent/non-intrusive standby mechanisms that reduce barriers
      of adoption.
%\end{enumerate}

Our proposal lies at the bottom of the software development
layers---programming language mechanisms---meaning that \emph{all} applications
should take advantage of low-power standby modes automatically, without extra
programming efforts.
%
We extend the programming language \CEU~\cite{ceu.sensys13,ceu.tecs17} with
support for interrupt service routines (ISRs) and with a simple power
management runtime (PMR).
%
In contrast with other concurrency models (e.g., thread and actor based), the
synchronous semantics of \CEU guarantees that reactions to the environment
always reach an idle state amenable to standby.
%
Previous work~\cite{ceu.sensys13} demonstrates the expressiveness of \CEU in
the context of Wireless Sensor Networks and discusses the development of
drivers, network protocols, and full applications in the language.
It also attests a small overhead of memory in comparison to C (around 5-10\%),
thus being a suitable alternative for constrained devices.
The language runtime is in the order of a few kilobytes only (less than 5Kb).

In our approach, each supported microcontroller requires hooks in C for the
ISRs and PMR, and each peripheral requires a driver in \CEU.
These are write-once code that is typically packaged and distributed in a
software development kit (SDK).
%
Then, all new applications built on top of these drivers take advantage of
standby automatically.
%
As a proof of concept, we provide an open source SDK %
%\footnote{https://github.com/fsantanna/ceu-arduino/}
with support for 8-bit
\emph{AVR/ATmega} and 32-bit \emph{ARM/Cortex-M0} microcontrollers, and a
variety of peripherals, such as for GPIO, A/D converter, USART, SPI, and the
nRF24L01 transceiver.
%
We developed a number of simple applications using these peripherals
concurrently and could verify that they remain in the deepest standby modes for
the longest periods of time.

@@

In Section~\ref{sec.ceu}, we compare the structure of programs in \CEU and
Arduino~\cite{arduino.book}, whose primary goal is to reduce the programming
barrier of adoption for a non-technical audience. % TODO-FULL (e.g., designers and artists).
We show that we can keep the intended sequential reasoning of Arduino even when
applications require non-trivial concurrent behavior.
%
In Section~\ref{sec.standby}, we discuss the software infrastructure that
allows for unmodified programs in \CEU to take advantage of standby
automatically.
%
In Section~\ref{sec.conclusion}, we discuss future work and conclude the paper.

@@

With all the standby infrastructure set, the unmodified program of
Figure~\ref{lst.inversion}.b automatically takes advantage of the deepest
sleep modes for the longest periods of time possible.

@@

In future work, we will evaluate the consumption of realistic applications.
%in order to evaluate the gains in energy efficiency with the
%proposed infrastructure,
%
The Arduino community has an abundance of open-source projects which can be
rewritten in \CEU to take advantage of transparent standby.
%
In this scenario, we can evaluate the time to rewrite, the resulting program
structure, and the actual energy efficiency.

\begin{acks}
The authors would like to thank Guilherme Simas for the initial explorations
with the language extensions and development of some of the device drivers.
This work was supported by the Serrapilheira Institute (grant number
Serra-1708-15612).
\end{acks}

\balance

### 2018-05-01 LCTES/lctes.tex
(>) LCTES: versão final

\usepackage{booktabs} % For formal tables
\usepackage{xspace}

\usepackage{listings}
\lstdefinelanguage{ceu}{%
  language=C,
  morekeywords={%
    and, uint32_t, code, async, isr,
    await, bool, call, class, data, define, deterministic, do, each, else,
    emit, end, escape, event, every, finalize, hor, implementation, in,
    input, interface, loop, min, native, new, nohold, not, or, output, par,
    pool, pure, return, signal, spawn, tag, then, this, traverse, until,
    var, watching, when, with},
}
\lstset{
  basicstyle=\ttfamily,
  captionpos=b,
  columns=flexible,
  commentstyle=\rmfamily\itshape,
  escapeinside={||},
  escapeinside={|},
  frame=tb,
  keepspaces=true,
  keywordstyle=\bfseries,
  language=ceu,
  mathescape=true,
  numbersep=4pt,
  numberstyle=\scriptsize,
  upquote=true,
}

@@

% Copyright
%\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
\acmDOI{10.475/123_4}

% ISBN
\acmISBN{123-4567-24-567/08/06}

%Conference
%\acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El Paso, Texas USA}
\acmConference[LCTES'18]{ACM SIGPLAN/SIGBED}{June 2018}{Philadelphia, USA}
%\acmYear{1997}
%\copyrightyear{2016}

%\acmPrice{15.00}

@@

@@

\begin{abstract}
Standby efficiency for connected devices is one of the priorities of the
\emph{G20's Energy Efficiency Action Plan}.
%
We propose transparent programming language mechanisms to enforce that
applications remain in the deepest standby modes for the longest periods of
time.
%
We extend the programming language \CEU with support for interrupt service
routines and with a simple power management runtime.
%
Based on these primitives, we also provide device drivers that allow
applications to take advantage of standby automatically.
%
Our approach relies on the synchronous semantics of the language which
guarantees that reactions to the environment always reach an idle state
amenable to standby.
%
In addition, to lower the programming barrier of adoption, we show that
programs in \CEU can keep a sequential syntactic structure, even when
applications require non-trivial concurrent behavior.
\end{abstract}

>>>>

% TODO-FULL
%\linespread{0.96}
\setlength{\belowcaptionskip}{-10pt}

\usepackage{booktabs} % For formal tables
\usepackage{xspace}
\usepackage{balance}

\usepackage{listings}
\lstdefinelanguage{ceu}{%
  language=C,
  morekeywords={%
    and, uint32_t, code, async, isr,
    await, bool, call, class, data, define, deterministic, do, each, else,
    emit, end, escape, event, every, finalize, hor, implementation, in,
    input, interface, loop, min, native, new, nohold, not, or, output, par,
    pool, pure, return, signal, spawn, tag, then, this, traverse, until,
    var, watching, when, with},
}
\lstset{
  basicstyle=\small\ttfamily,
  captionpos=b,
  columns=flexible,
  commentstyle=\rmfamily\itshape,
  escapeinside={||},
  escapeinside={|},
  frame=tb,
  keepspaces=true,
  keywordstyle=\bfseries,
  language=ceu,
  mathescape=true,
  numbersep=4pt,
  numberstyle=\scriptsize,
  upquote=true,
}

@@

% Copyright
%\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

@@

%%% If you see 'ACMUNKNOWN' in the 'setcopyright' statement below,
%%% please first submit your publishing-rights agreement with ACM (follow link on submission page).
%%% Then please update our instructions page and copy-and-paste the NEW commands into your article.
%%% Please contact us in case of questions; allow up to 10 min for the system to propagate the information.
%%%
%%% The following is specific to LCTES'18 and the paper
%%% 'Transparent Standby for Low-Power, Resource-Constrained Embedded Systems: A Programming Language-Based Approach (Short WIP Paper)'
%%% by Francisco Sant'Anna, Alexandre Sztajnberg, and Noemi Rodrigues.
%%%
%\setcopyright{ACMUNKNOWN}
\setcopyright{acmcopyright}
\acmPrice{15.00}
\acmDOI{10.1145/3211332.3211337}
\acmYear{2018}
\copyrightyear{2018}
\acmISBN{978-1-4503-5803-3/18/06}
\acmConference[LCTES'18]{19th ACM SIGPLAN/SIGBED Conference on Languages, Compilers, and Tools for Embedded Systems}{June 19--20, 2018}{Philadelphia, PA, USA}

@@

\begin{abstract}
Standby efficiency for connected devices is one of the priorities of the
\emph{G20's Energy Efficiency Action Plan}.
%
We propose transparent programming language mechanisms to enforce that
applications remain in the deepest standby modes for the longest periods of
time.
%
We extend the programming language \CEU with support for interrupt service
routines and with a simple power management runtime.
%
Based on these primitives, we also provide device drivers that allow
applications to take advantage of standby automatically.
%
Our approach relies on the synchronous semantics of the language which
guarantees that reactions to the environment always reach an idle state
amenable to standby.
%
In addition, in order to lower the programming barrier of adoption, we show
that programs in \CEU can keep a sequential syntactic structure, even when
applications require non-trivial concurrent behavior.
\end{abstract}
